Extended Early Adulthood and Attitude Change

### 2024-04-08 manuscript/formative_period_kk.Rmd
new_approach

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "7/06/2023"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: Young adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture. This probability has often been attributed to the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage and childbirth have transformed and extended this life stage. We leverage these changes to explore whether life course changes, as well as broader social changes, have affected how likely people of different cohorts are to make durable changes of general political dispositions over a comparable time frame. We find strikingly little difference in the rates of any kind of stable opinions and opinion change among young people in different eras. People of all ages have similar attitude behaviors over time. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates. 
---

@@

However, why people in this life stage appear to show more frequent opinion change relative to older people is unclear. Explanations in the social sciences tend to highlight the unique social features of these life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information. Work in biology, psychology, and cultural evolution, in contrast, shows that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION]. While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.

Clarifying the relative importance of different mechanisms that drive the frequency of opinion change during this formative period is important for understanding the mutually reinforcing processes of social change and cultural evolution. If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time. If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social change in the same way.

In this paper, we explore how changes in the broader social structure and changes in the timing of life course events since the 1950s have affected the stability and reliability of political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population. Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005]. Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. This suggests both a lengthening of and an intensification of this stage of the life course in ways that potentially affect the rates at which people undergo cultural change.

We take advantage of this change over time to explore whether young adults of different cohorts demonstrate more or less change in their political opinions over time relative to other members of the population. We decompose over-time responses into reliability (a lack of random, non-persisting fluctuations) and stability (true consistency in latent dispositions net of these random fluctuations) for a range of political attitudes in nine nationally representative panel data sets spanning from the 1950s through the 2010s. We separately estimate these parameters for a comparable set of age groups in each panel to evaluate whether response reliability or stability changed as new cohorts moved through the same age ranges under different social conditions.

@@

@@

We divide each sample into seven age categories: 18-25, 26-33, 34-41, 42-49, 50-57, 58-65, 66-83, assessed at the initial survey year. We analyze these estimates in two ways. First, we explore the stability of the latent attitudes, $\beta_21$ and $\beta_32$. These estimates are non-independent, so we examine them as multiple indicators of the stability of the attitude in that panel. We take these measures of stability as a measure of how much durable change in attitudes occurs in an issue during the panel, with higher values indicating less change. Second, we examine a single reliability estimate generated for each group for each attitude measure. We take this as a combined measure of actual measurement error and transitive, non-durable change that respondents make during the course of the panel. For each age group, we average across all questions in a given panel to generate an average estimate of reliability and an average estimate of stability for each age group in each panel.

We compare these average estimates of reliability and stability for age groups across panels to evaluate our hypotheses.

For questions that are asked in multiple panels, we also use estimates of stability and reliability as outcome variables in a regression on time with fixed effects for questions. We run these models separately for each age group. These models let us allow for an average trajectory of reliability and stability for age groups over time, controlling for questions, to evaluate whether respondents in the same age groups have become more or less stable reporters of the same opinions over time. We also control for attributes of panels that might affect reliability or stability estimates -- such as the shortened wave duration in the 1980 panel or the mode of administration -- by including panel fixed effects in the model as well.

>>>>

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: Early adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture. This probability has often been attributed to the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage and childbirth have transformed and extended this life stage. We leverage these changes to explore whether life course changes, as well as broader social changes, have affected how likely people of different cohorts are to make durable changes of general political dispositions over a comparable time frame. We find strikingly little difference in the rates of opinion change among young people in different eras. People of all ages have similar attitude behaviors over time. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates of change. 
---

@@

However, why people in this life stage appear to show more frequent opinion change relative to older people is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Alternative explanations in the social sciences also focus on the unique features of middle age that tend to stabilize attitudes during this period, including higher levels of social authority [@eaton2009; @visser2004]. Work in biology, psychology, and cultural evolution, in contrast, shows that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION]. While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.

Clarifying the relative importance of different mechanisms that drive the frequency of opinion change during this formative period is important for understanding the mutually reinforcing processes of social change and cultural evolution. If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016]. If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social change in the same way.

In this paper, we explore how rates of opinion change in different age groups have changed in response to broader social structure and changes in the timing of life course events since the 1950s. Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005]. Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. This suggests both a lengthening of and an intensification of this stage of the life course in ways that potentially affect the rates at which people undergo cultural change.

We take advantage of this change over time to explore whether young adults of different cohorts demonstrate more or less change in their political opinions over time relative to other members of the population.

We decompose over-time responses into reliability (a lack of random, non-persisting fluctuations) and stability (true consistency in latent dispositions net of these random fluctuations) for a range of political attitudes in nine nationally representative panel data sets spanning from the 1950s through the 2010s. We separately estimate these parameters for a comparable set of age groups in each panel to evaluate whether response reliability or stability changed as new cohorts moved through the same age ranges under different social conditions.

@@

There are a broad range of approaches to measure the amount of change a group undergoes. We could compare the average slope of different age groups, but this assumes that , and would miss

My starting point for our analysis if the Life Course Adaption Model (LCAM) formalized by Lersch [-@lersch2023]. The LCAM is a mixed-effects growth curve model that formalizes several key insights from the literature on attitude change over time. First, it assumes that survey respondents have an underlying disposition shaped by early-life socialization. Second, it assumes that people can, but do not necessarily, change this disposition over time as a result of aging and general life-course experiences. Third, it assumes that people fluctuate around their underlying dispositions over time as a result of random shocks and measurement error. Fourth, it assumes respondents can draw on their prior experiences when adapting to new events.

The model takes the form of:

$$ c_{i,t}^* = \beta_0 + \beta_ia_{i,t} + u_{0,i} + u_{i,t}a_{i,t} + \epsilon_{i,t}$$

where $c_{i,t}^*$ is the standardized survey response for respondent $i$ at time $t$, $\beta_0$ is the intercept with random term $u_{0,i}$ capturing early-life experiences. $\beta_1$ with random term $u_{i,t}a_{i,t}$ captures the individual-specific slope for age $a_{i,t}$. $\epsilon_{i,t}$ is random error around this individual-specific slope for person $i$ at time $t$.

Because we are unconcerned with whether respondents change in the same direction or tend to change in different directions (which could indicate either convergence or polarization, depending on where respondents start). The central quantity of interest for our analysis is the absolute change in responses for a year, or $|\beta_ia_{i,t} + u_{i,t}a_{i,t}|$. Averaging this quantity across all respondents in a group produces an average absolute change in responses for a year for members of that group.

As a second quantity of interest, we calculate the mean square residual for respondents in a particular group. Groups might differ both in their linear (directional) change, as well as whether they are consistent at respodning around that trajectory. High variance around a latent trajectory could indicate that respondents are either more error-prone in making responses or more subject to temporary considerations. To calculate this, we calculate:

$$\frac{1}{n}\sum{(y_{i,t} - \hat{y_{i,t}})^2} $$

Our central question deals with comparisons across age groups at one time and within age groups over time in the amount of change we observe. Because of this, we need a way to extract a single measure of change .

@@
